
# Loading our data {#loading-data}
If you are using the VM, you will already have available some data to play around with. However, we are going to use a different dataset for most of our analysis. Kaggle hosts a dataset containing data from an ecommerce-site, which is both interesting and quite conductive to learning advanced SQL.

In order to download the data you need a Kaggle account, but once that is set up, you can download the data from https://www.kaggle.com/carrie1/ecommerce-data/downloads/ecommerce-data.zip.

In order to load it into the SQL database, we will first load it into R as a regular dataframe and then write it to a table.

```{r dbdrvx, include=FALSE, eval=TRUE}

library(ROracle)
drv <- dbDriver("Oracle")
host <- "localhost"
port <- 1521
sid <- "orcl"

connect.string <- paste(
  "(DESCRIPTION=",
  "(ADDRESS=(PROTOCOL=tcp)(HOST=", host, ")(PORT=", port, "))",
  "(CONNECT_DATA=(SID=", sid, ")))", sep = "")

con <- dbConnect(drv, username="moviedemo", password="welcome1", dbname=connect.string)
tickercon <- dbConnect(drv, username="pmuser", password="pmuser", dbname=connect.string)

```


```{r readcsv}
trx <- read.csv("data/online_transactions.csv", stringsAsFactors = F)
trx$InvoiceDate <- strptime(x = trx$InvoiceDate, format = "%m/%d/%Y %H:%M")
names(trx) <- lapply(names(trx), toupper)
```

The `InvoiceDate` is a timestamp but R reads it as a string instead of a date, so we needed to convert it with the `strptime` command. We also convert all variable names to uppercase, as Oracle seems strangely sensitive to this (SQL is not usually case-sensitive).

Writing the dataframe to a new table in the `moviedemo` database can be done with `dbWriteTable`, a function in the `ROracle` library that needs only three arguments: the database connection object, a new table name, and the name of our dataframe.

```{r insertdata, include=TRUE, eval=FALSE}
dbWriteTable(connection, "TRX0", trx)
```

Unfortunately, the date format is lost from R to Oracle so we need to do a little data transformation in SQL as well. We do want the date field as an actual date, even though dates can be a hassle (like now), they prove very valuable for anyone who invests some time in learning about them.

This statement could have been executed from R, but using the SQL Workbench is probably just as nice.

```sql
CREATE TABLE trx AS (
SELECT INVOICENO, 
  STOCKCODE, 
  DESCRIPTION, 
  QUANTITY, 
  TO_DATE(INVOICEDATE, 'YYYY-MM-DD HH24:MI:SS') AS INVOICEDATE, 
  UNITPRICE, 
  CUSTOMERID,
  COUNTRY
FROM moviedemo.trx0
)
```
